Fail Faster

### 2025-10-06 paper/paper.tex
paper

We evaluate the optimizations with a series of case studies in each
generator library, and show that the \name{} generators run faster
than their unstaged equivalents. In \camlname, they run up to $7\times$
faster, and in \scalaname, up to $13\times$ faster. Using our improved version
of \bq's \rand in \camlname, we see even greater gains---more than $12\times$---showing that
both staging and fast randomness play distinct and complementary roles in improving generator performance.
Further,
we show that \camlname generators find bugs faster by running our case studies
in the Etna platform, which uses generated values as
inputs to a buggy system and measures how quickly different
generators can detect the bugs. In Etna,
\camlname generators find bugs up to $2.5\times$ faster on average than
their \bq counterparts. With fast randomness,
the average speedup rises to $3.8\times$,
with some cases exceeding $13\times$.

In summary, we show that PBT generator DSLs incur significant
performance costs across languages and present two interventions that
significantly improve their efficiency while preserving their
idiomatic style.  Concretely, we offer the following contributions:
\begin{enumerate}
\item We identify two key sources of inefficiency in PBT generator
libraries---abstraction overhead and choice of \rand---both of which
significantly impact performance (Section~\ref{section:motiv}).
\item We present \name{}, a staging technique that eliminates the
abstraction overhead of generators. We apply \name{} to standard
generator libraries in both OCaml and Scala 3, showcasing its
generality (Section~\ref{section:faster-generators}).
\item We demonstrate that writing generators using \name{} and fast
\rands{} yields substantial performance improvements, both separately
and in combination, and we show that these performance improvements
extend to significantly improved bug-finding speed (Section~\ref{section:eval}).
\end{enumerate}

Section~\ref{section:other-langs} discusses how the \name{}
technique could be applied to PBT libraries in other
languages---Racket, F\#, Haskell, and Rust.
%
Sections~\ref{section:related} and~\ref{section:concl} present related
and future work.

@@

\section{Beyond OCaml and Scala}
\label{section:other-langs}

>>>>

We evaluate the optimizations with a series of case studies in each
generator library, and show that the \name{} generators run faster
than their unstaged equivalents. In \camlname, they run up to $7\times$
faster, and in \scalaname, up to $13\times$ faster. Using our improved version
of \bq's \rand in \camlname, we see even greater gains---more than $12\times$---showing that
both staging and fast randomness play distinct and complementary roles in improving generator performance.
Further,
we show that \camlname generators find bugs faster by running our case studies
in the Etna platform, which uses generated values as
inputs to a buggy system and measures how quickly different
generators can detect the bugs. In Etna,
\camlname generators find bugs up to $2.5\times$ faster on average than
their \bq counterparts. With fast randomness,
the average speedup rises to $3.8\times$,
with some cases exceeding $13\times$. Finally, we review related libraries
in other languages including Racket, F\#, Haskell, Rust, Clojure, and Swift,
and discuss how the \name{} technique could be applied to them.

In summary, we show that PBT generator DSLs incur significant
performance costs across languages and present two interventions that
significantly improve their efficiency while preserving their
idiomatic style.  Concretely, we offer the following contributions:
\begin{enumerate}
\item We identify two key sources of inefficiency in PBT generator
libraries---abstraction overhead and choice of \rand---both of which
significantly impact performance (Section~\ref{section:motiv}).
\item We present \name{}, a staging technique that eliminates the
abstraction overhead of generators. We apply \name{} to standard
generator libraries in both OCaml and Scala 3, showcasing its
generality (Section~\ref{section:faster-generators}).
\item We present a novel methodology for quantifying 
      the precise speedup achieved using staging and fast randomness,
      in which the performance benefit of each intervention can be
      measured in isolation with absolutely no noise (Section~\ref{section:eval}).
\item We demonstrate that writing generators using \name{} and fast
\rands{} yields substantial performance improvements, both separately
and in combination, and we show that these performance improvements
extend to significantly improved bug-finding speed (Section~\ref{section:eval}).
\end{enumerate}

Section~\ref{section:other-langs} discusses how the \name{}
technique could be applied to PBT libraries in other
languages---Racket, F\#, Haskell, Rust, Clojure, and Swift.
%
Sections~\ref{section:related} and~\ref{section:concl} present related
and future work.

@@

\section{Other Languages \& Libraries}
\label{section:other-langs}
